---
title: "Local Linear Regression for Estimating Monotonic Biological Rates in R (LoLinR)"
author: "Colin Olito and Diego Barneche"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, results="hide", warnings=FALSE, echo=FALSE, message=FALSE}
devtools::load_all()
```

In this `html_vignette`, we introduce the `LoLinR` package, which provides tools to implement local linear regression techniques for estimating monotonic biological rates from time-series or trace data in a statistically robust and reproducible fashion. The methods are a modification of traditional Loess regression techniques built around the wrapper function `rankLocReg()`. 

**1. Motivation**  

Extract from the paper. Colin's new change.

**2. Sea urchin data**  

We here use respiration data from the sea urchin *Heliocidaris erythrogramma*, common to Port Phillip Bay, Melbourne, Australia. The data are *O~2~* consumption time-series for 4 individual urchins included in a respirometry study (Colin Olito *unpublished data*). For this example, the biological question of interest has to do with estimating resting metabolic rate for each individual from their *O~2~* consumption time series. The variable include Time (*min*), and *mL O~2~* data for each individual (coded by the letters A-D). After loading the \code{UrchinData} data set, examination of the time series for individual 'D' gives a representative example of the analytical challenge presented by these data.

```{r, fig.width=5, fig.height=5}  
data(UrchinData)
par(omi=rep(0.3, 4))
plot(D ~ time, data=UrchinData, 
    xlab='Time (minutes)', ylab=substitute('Volume O'[2]~' (mL)'), type='n', axes=FALSE)
usr  <-  par('usr')
rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
whiteGrid()
box()
points(UrchinData$D ~ UrchinData$time, pch=16, col=transparentColor('dodgerblue2', 0.6), cex=1.1)
axis(1)
axis(2, las=1)
```
  
The full data set is clearly non-linear. As is typical of closed-chamber respirometry data, the initial 20 min of the data are messy, as the animal is stressed from handling, and the chamber equilibrates. There is also a more subtle acceleration in the rate of *O~2~* consumption towards the end of the time series as the animal exhibits a physiological response to declining *O~2~* avaiability in the chamber. While a non-linear statistical model could conceivably be fit to these data, the question of interest requires an estimate of a simple monotonic rate. Although estimating a simple monotonic rate from these data appears to be straightforward, it is actually a non-trivial challenge. On one hand, any estimate that includes the non-linear portions of the data is not actually an estimate of the rate of interest, as it includes physiological responses to stress and declining oxygen availability. On the other hand, truncating the data to exclude the non-linear parts is subjective, can sacrifice statistical power by excluding good data, and may be arbitrary, not reproducible, or both. Ultimately, we are interested in the region where the rate of *O~2~* consumption is most stable, which in theory should represent the resting metabolic rate. Using \pkg{LoLinR}, we can do this using the \code{rankLocReg} function.

```{r}
urchinRegs  <-  rankLocReg(xall=UrchinData$time, yall=UrchinData$D, alpha=0.2, 
    method="eq", verbose=TRUE)  
```

The main function `rankLocReg` runs all possible local regressions (on adjacent data points) with minimum window size `alpha`. It chooses one of three methods/metrics (`ns`, `eq` or `pc` -- default to `ns`, see below for explanation of each method) to rank the best local regression. It outputs an object of class `rankLocReg`  

*Explain different methods*

```{r}
summary(urchinRegs)
```

*Explain what each output is here*

Plot output for object of class `rankLocReg`, using the best local regression (according to method `eq`):  

```{r, eval=FALSE}
plot(urchinRegs, rank=1)
```

```{r, echo=FALSE, fig.width=9, fig.height=5}
    x     <-  urchinRegs
    rank  <-  1
    #  recover data window for chosen local regression model
    bestwin  <-  c(x$allRegs$Lbound[rank], x$allRegs$Rbound[rank])
    y1       <-  x$yall[bestwin[1]:bestwin[2]]
    x1       <-  x$xall[bestwin[1]:bestwin[2]]
    
    #  fit block
    fit     <-  locReg(bestwin, x$xall, x$yall, resids=TRUE)
    locFit  <-  fit$table
    resids  <-  fit$residuals
    b1      <-  locFit$b1
    yHat    <-  fit$yHat

    #  residual plots
    layout(matrix(c(
                    rep(c(rep(1, 4), rep(2, 2), rep(3, 2)), 2),
                    rep(c(rep(1, 4), rep(4, 2), rep(5, 2)), 2)
                   ), 
           nrow=4, ncol=8, byrow=TRUE)
    )
    
    #  overall regression plot
    outy  <-  x$yall[c(1:(bestwin[1]-1), (bestwin[2]+1):length(x$yall))]
    outx  <-  x$xall[c(1:(bestwin[1]-1), (bestwin[2]+1):length(x$yall))]

    par(mai=c(1.2, 0.8, 0.8, 0.4), cex=1)
    plot(x$yall ~ x$xall, axes=FALSE, type='n', xlab='Predictor', ylab='Response', cex.lab=1.2, ylim=c(min(x$yall), (max(x$yall) + 0.1*(max(x$yall) - min(x$yall)))))
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    points(outy ~ outx, pch=16, col=transparentColor('black', 0.2), cex=1.2)
    points(y1 ~ x1, col='dodgerblue', cex=1.2)
    lines(x1, locFit$b0 + locFit$b1*x1, col='black', lwd=2, lty=2)
    proportionalLabel(c(0, 0.14), rep(1.1, 2), text=FALSE, xpd=NA, type='l', lwd=2, lty=2)
    proportionalLabel(0.15, 1.1, substitute('Rank '*pos*': '*italic(y) == a~sy~b%.%italic(x), list(pos=rank, a=rounded(locFit$b0, 2), sy=ifelse(b1 < 0, ' - ', ' + '), b=rounded(abs(b1), 2))), xpd=NA, adj=c(0, 0.5))
    proportionalLabel(c(0, 0.14), rep(1.1, 2), text=FALSE, xpd=NA, type='l', lwd=2, lty=2)
    proportionalLabel(0.95, 0.95, paste0('n = ', length(y1)), xpd=NA, adj=c(1, 0.5), font=3, col='dodgerblue')

    # standardized residuals ~ x
    par(mai=c(0.6732, 0.5412, 0.5412, 0.2772), cex=0.8)
    yRange  <-  max(abs(c(floor(min(resids)), ceiling(max(resids)))))
    yRange  <-  c(-1*yRange, yRange)
    plot(resids ~ x1, xlab='Predictor', ylab='Std. residuals', xpd=NA, ylim=yRange, type='n', axes=FALSE)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    points(resids ~ x1, pch=16, col=transparentColor('dodgerblue', 0.5))
    abline(h=0, col=1, lwd=2)
    abline(h=c(-2, 2), lty=2)
    lf1  <-  loess(resids ~ x1)
    lines(x1, lf1$fitted, col='tomato', lwd=2)
    
    # standardized residuals ~ fitted values
    plot(resids ~ yHat, xlab='Fitted Values', ylab='Std. residuals', xpd=NA, ylim=yRange, type='n', axes=FALSE)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    points(resids ~ yHat, pch=16, col=transparentColor('dodgerblue', 0.5))
    abline(h=0, col=1, lwd=2)
    abline(h=c(-2, 2), lty=2)
    lf2  <-  loess(resids ~ yHat)
    lines(yHat, lf2$fitted, col='tomato', lwd=2)
    
    # qqnorm plot of standardized residuals
    par(mai=c(0.9732, 0.5412, 0.2412, 0.2772), cex=0.8)
    qqPlot  <-  qqnorm(resids, main='QQNorm plot of Std. Residuals', xpd=NA, plot=FALSE)
    plot(y1 ~ x1, data=qqPlot, xlab='Theoretical quantiles', ylab='Sample quantiles', xpd=NA, ylim=yRange, xlim=yRange, type='n', axes=FALSE)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    points(qqPlot$y ~ qqPlot$x, pch=16, col=transparentColor('dodgerblue', 0.5))
    qqline(resids, col='tomato')
    
    # histogram of standardized residuals
    histPlot  <-  hist(resids, breaks=20, plot=FALSE)
    plot(NA, xlab='Std. Residuals', ylab='Density', xpd=NA, ylim=c(0, max(histPlot$density)), xlim=yRange, type='n', axes=FALSE)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    densities  <-  histPlot$density
    breakPts   <-  histPlot$breaks
    for(j in seq_along(densities)) {
        polygon(c(breakPts[j], breakPts[j+1], breakPts[j+1], breakPts[j], breakPts[j]), c(rep(usr[3], 2), rep(densities[j], 2), usr[3]), border='dodgerblue', col=transparentColor('dodgerblue', 0.5))
    }
```
One can also inspect the best 25 local regressions following the chosen method  

```{r, eval=FALSE}
outputRankLocRegPlot(urchinRegs)
```

```{r, echo=FALSE, fig.width=7, fig.height=7}
	allRegs  <-  urchinRegs
    par(mfrow=c(5,5), omi=rep(1, 4), mai=rep(0,4), cex=1)
    locFit  <-  allRegs$allRegs

    for(i in 1:25) {
        # subset data
        outy  <-  allRegs$yall[c(1:(locFit$Lbound[i]-1), (locFit$Rbound[i]+1):length(allRegs$yall))]
        outx  <-  allRegs$xall[c(1:(locFit$Lbound[i]-1), (locFit$Rbound[i]+1):length(allRegs$yall))]
        y     <-  allRegs$yall[locFit$Lbound[i]:locFit$Rbound[i]]
        x     <-  allRegs$xall[locFit$Lbound[i]:locFit$Rbound[i]]

        # plot
        plot(allRegs$yall ~ allRegs$xall, axes=FALSE, type='n', xlab='Predictor', ylab='Response', cex.lab=1.2, ylim=c(min(allRegs$yall), (max(allRegs$yall) + 0.25*(max(allRegs$yall) - min(allRegs$yall)))))
        usr  <-  par('usr')
        rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
        whiteGrid()
        box()

        # check whether axes and labels are to be plotted
        if(i %in% seq(1, 21, 5))
            axis(2, las=1, cex.axis=0.5)
        if(i %in% 21:25)
            axis(1, cex.axis=0.5, mgp=c(3, 0.5, 0))
        
        points(outy ~ outx, pch=16, col=transparentColor('black', 0.2), cex=1.2)
        points(y ~ x, col='dodgerblue', cex=0.8)
        lines(x, locFit$b0[i] + locFit$b1[i]*x, col='black', lwd=2, lty=2)
        proportionalLabel(0.03, 0.9, substitute(italic(z)*italic(y) == a~sy~b%.%italic(x), list(z=paste0(i, ';   '), a=rounded(locFit$b0[i], 2), sy=ifelse(locFit$b1[i] < 0, ' - ', ' + '), b=rounded(abs(locFit$b1[i]), 2))), adj=c(0, 0.5), cex=0.5)
    }
    mtext('Response', side=2, line=2.5, outer=TRUE)
    mtext('Predictor', side=1, line=2.5, outer=TRUE)
```

Or even check the estimated slope for the best local regression for each of the ranking methods benchmarked against a distribution of all estimated slopes

```{r, eval=FALSE}
plotBeta1(urchinRegs)
```

```{r, echo=FALSE, fig.width=5, fig.height=5}
	allRegs  <-  urchinRegs

    c1  <-  'tomato'
    c2  <-  'darkolivegreen'
    c3  <-  'dodgerblue4'
    
    par(omi=rep(0.5, 4), cex=1)
    locFit     <-  allRegs$allRegs
    b1Density  <-  density(locFit$b1)

    plot(NA, xlab=expression(paste(beta[1])), type='n', axes=FALSE, ylab='Density', cex.lab=1.2, xlim=range(b1Density$x), ylim=c(0, (max(b1Density$y)+0.05*max(b1Density$y))), yaxs='i')
    proportionalLabel(0.5, 1.1, expression(paste('Distribution of ', beta[1])), xpd=NA, adj=c(0.5, 0.5), font=3, cex=1.4)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    polygon(c(b1Density$x), c(b1Density$y), col=transparentColor('dodgerblue2', 0.5), border='dodgerblue2')
    axis(1)
    axis(2, las=1)

    abline(v=locFit$b1[locFit$L == min(locFit$L)], col=c1, lty=1, lwd=3)
    abline(v=locFit$b1[locFit$Leq == min(locFit$Leq)], col=c2, lty=2, lwd=3)
    abline(v=locFit$b1[locFit$Lpc == min(locFit$Lpc)], col=c3, lty=3, lwd=3)
    legend(
          x       =  usr[2],
          y       =  usr[4],
          legend  =  c(expression(paste(italic(L))),
                      expression(paste(italic(L[eq]))),
                      expression(paste(italic(L['%'])))),
          lwd     =  4,
          lty     =  c(1, 2, 3),
          col     =  c(c1, c2, c3),
          cex     =  1,
          xjust   =  1,
          yjust   =  1,
          bty     =  'n',
          border  =  NA
    )
```

**3. O~2~ consumption in cormorant**  

We now show a different utility using a dataset containing O~2~ consumption data for a Great cormorant. This dataset represents a small example of data used for analyses described in White et al. 2011. Metabolic rate throughout the annual cycle reveals the demands of an Arctic existence in Great Cormorants. Ecology 92: 475--486.  

Here we thin this dataset to speed up computation time using the package function `thinData`. *Explain the rationale better*.

*Describe why this flow-through dataset is different*. Using method `eq`:

```{r}
data(CormorantData)
CormorantData  <-  thinData(CormorantData, by=3)$newData1
cormRegs  <-  rankLocReg(xall=CormorantData$Time, yall=CormorantData$VO2.ml.min, alpha=0.2, 
    method="eq", verbose=FALSE)
```

As we can see, the best local regression picked up by method `eq` is not ideal.

```{r, eval=FALSE}
plot(cormRegs, rank=1)
```

```{r, echo=FALSE, fig.width=9, fig.height=5}
    x     <-  cormRegs
    rank  <-  1
    #  recover data window for chosen local regression model
    bestwin  <-  c(x$allRegs$Lbound[rank], x$allRegs$Rbound[rank])
    y1       <-  x$yall[bestwin[1]:bestwin[2]]
    x1       <-  x$xall[bestwin[1]:bestwin[2]]
    
    #  fit block
    fit     <-  locReg(bestwin, x$xall, x$yall, resids=TRUE)
    locFit  <-  fit$table
    resids  <-  fit$residuals
    b1      <-  locFit$b1
    yHat    <-  fit$yHat

    #  residual plots
    layout(matrix(c(
                    rep(c(rep(1, 4), rep(2, 2), rep(3, 2)), 2),
                    rep(c(rep(1, 4), rep(4, 2), rep(5, 2)), 2)
                   ), 
           nrow=4, ncol=8, byrow=TRUE)
    )
    
    #  overall regression plot
    outy  <-  x$yall[c(1:(bestwin[1]-1), (bestwin[2]+1):length(x$yall))]
    outx  <-  x$xall[c(1:(bestwin[1]-1), (bestwin[2]+1):length(x$yall))]

    par(mai=c(1.2, 0.8, 0.8, 0.4), cex=1)
    plot(x$yall ~ x$xall, axes=FALSE, type='n', xlab='Predictor', ylab='Response', cex.lab=1.2, ylim=c(min(x$yall), (max(x$yall) + 0.1*(max(x$yall) - min(x$yall)))))
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    points(outy ~ outx, pch=16, col=transparentColor('black', 0.2), cex=1.2)
    points(y1 ~ x1, col='dodgerblue', cex=1.2)
    lines(x1, locFit$b0 + locFit$b1*x1, col='black', lwd=2, lty=2)
    proportionalLabel(c(0, 0.14), rep(1.1, 2), text=FALSE, xpd=NA, type='l', lwd=2, lty=2)
    proportionalLabel(0.15, 1.1, substitute('Rank '*pos*': '*italic(y) == a~sy~b%.%italic(x), list(pos=rank, a=rounded(locFit$b0, 2), sy=ifelse(b1 < 0, ' - ', ' + '), b=rounded(abs(b1), 2))), xpd=NA, adj=c(0, 0.5))
    proportionalLabel(c(0, 0.14), rep(1.1, 2), text=FALSE, xpd=NA, type='l', lwd=2, lty=2)
    proportionalLabel(0.95, 0.95, paste0('n = ', length(y1)), xpd=NA, adj=c(1, 0.5), font=3, col='dodgerblue')

    # standardized residuals ~ x
    par(mai=c(0.6732, 0.5412, 0.5412, 0.2772), cex=0.8)
    yRange  <-  max(abs(c(floor(min(resids)), ceiling(max(resids)))))
    yRange  <-  c(-1*yRange, yRange)
    plot(resids ~ x1, xlab='Predictor', ylab='Std. residuals', xpd=NA, ylim=yRange, type='n', axes=FALSE)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    points(resids ~ x1, pch=16, col=transparentColor('dodgerblue', 0.5))
    abline(h=0, col=1, lwd=2)
    abline(h=c(-2, 2), lty=2)
    lf1  <-  loess(resids ~ x1)
    lines(x1, lf1$fitted, col='tomato', lwd=2)
    
    # standardized residuals ~ fitted values
    plot(resids ~ yHat, xlab='Fitted Values', ylab='Std. residuals', xpd=NA, ylim=yRange, type='n', axes=FALSE)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    points(resids ~ yHat, pch=16, col=transparentColor('dodgerblue', 0.5))
    abline(h=0, col=1, lwd=2)
    abline(h=c(-2, 2), lty=2)
    lf2  <-  loess(resids ~ yHat)
    lines(yHat, lf2$fitted, col='tomato', lwd=2)
    
    # qqnorm plot of standardized residuals
    par(mai=c(0.9732, 0.5412, 0.2412, 0.2772), cex=0.8)
    qqPlot  <-  qqnorm(resids, main='QQNorm plot of Std. Residuals', xpd=NA, plot=FALSE)
    plot(y1 ~ x1, data=qqPlot, xlab='Theoretical quantiles', ylab='Sample quantiles', xpd=NA, ylim=yRange, xlim=yRange, type='n', axes=FALSE)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    points(qqPlot$y ~ qqPlot$x, pch=16, col=transparentColor('dodgerblue', 0.5))
    qqline(resids, col='tomato')
    
    # histogram of standardized residuals
    histPlot  <-  hist(resids, breaks=20, plot=FALSE)
    plot(NA, xlab='Std. Residuals', ylab='Density', xpd=NA, ylim=c(0, max(histPlot$density)), xlim=yRange, type='n', axes=FALSE)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    densities  <-  histPlot$density
    breakPts   <-  histPlot$breaks
    for(j in seq_along(densities)) {
        polygon(c(breakPts[j], breakPts[j+1], breakPts[j+1], breakPts[j], breakPts[j]), c(rep(usr[3], 2), rep(densities[j], 2), usr[3]), border='dodgerblue', col=transparentColor('dodgerblue', 0.5))
    }
```

When inspecting the distribution of all local regression slopes, we see that the percentage method might be better  

```{r, eval=FALSE}
plotBeta1(cormRegs)
```

```{r, echo=FALSE, fig.width=5, fig.height=5}
	allRegs  <-  cormRegs

    c1  <-  'tomato'
    c2  <-  'darkolivegreen'
    c3  <-  'dodgerblue4'
    
    par(omi=rep(0.5, 4), cex=1)
    locFit     <-  allRegs$allRegs
    b1Density  <-  density(locFit$b1)

    plot(NA, xlab=expression(paste(beta[1])), type='n', axes=FALSE, ylab='Density', cex.lab=1.2, xlim=range(b1Density$x), ylim=c(0, (max(b1Density$y)+0.05*max(b1Density$y))), yaxs='i')
    proportionalLabel(0.5, 1.1, expression(paste('Distribution of ', beta[1])), xpd=NA, adj=c(0.5, 0.5), font=3, cex=1.4)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    polygon(c(b1Density$x), c(b1Density$y), col=transparentColor('dodgerblue2', 0.5), border='dodgerblue2')
    axis(1)
    axis(2, las=1)

    abline(v=locFit$b1[locFit$L == min(locFit$L)], col=c1, lty=1, lwd=3)
    abline(v=locFit$b1[locFit$Leq == min(locFit$Leq)], col=c2, lty=2, lwd=3)
    abline(v=locFit$b1[locFit$Lpc == min(locFit$Lpc)], col=c3, lty=3, lwd=3)
    legend(
          x       =  usr[2],
          y       =  usr[4],
          legend  =  c(expression(paste(italic(L))),
                      expression(paste(italic(L[eq]))),
                      expression(paste(italic(L['%'])))),
          lwd     =  4,
          lty     =  c(1, 2, 3),
          col     =  c(c1, c2, c3),
          cex     =  1,
          xjust   =  1,
          yjust   =  1,
          bty     =  'n',
          border  =  NA
    )
```

So we can easily re-sort the output `data.frame` of local regressions using method `pc`

```{r, eval=FALSE}
cormRegs  <-  reRank(cormRegs, newMethod='pc')
plot(cormRegs, rank=1)
```
```{r, echo=FALSE, fig.width=9, fig.height=5}
    x     <-  reRank(cormRegs, newMethod='pc')
    rank  <-  1
    #  recover data window for chosen local regression model
    bestwin  <-  c(x$allRegs$Lbound[rank], x$allRegs$Rbound[rank])
    y1       <-  x$yall[bestwin[1]:bestwin[2]]
    x1       <-  x$xall[bestwin[1]:bestwin[2]]
    
    #  fit block
    fit     <-  locReg(bestwin, x$xall, x$yall, resids=TRUE)
    locFit  <-  fit$table
    resids  <-  fit$residuals
    b1      <-  locFit$b1
    yHat    <-  fit$yHat

    #  residual plots
    layout(matrix(c(
                    rep(c(rep(1, 4), rep(2, 2), rep(3, 2)), 2),
                    rep(c(rep(1, 4), rep(4, 2), rep(5, 2)), 2)
                   ), 
           nrow=4, ncol=8, byrow=TRUE)
    )
    
    #  overall regression plot
    outy  <-  x$yall[c(1:(bestwin[1]-1), (bestwin[2]+1):length(x$yall))]
    outx  <-  x$xall[c(1:(bestwin[1]-1), (bestwin[2]+1):length(x$yall))]

    par(mai=c(1.2, 0.8, 0.8, 0.4), cex=1)
    plot(x$yall ~ x$xall, axes=FALSE, type='n', xlab='Predictor', ylab='Response', cex.lab=1.2, ylim=c(min(x$yall), (max(x$yall) + 0.1*(max(x$yall) - min(x$yall)))))
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    points(outy ~ outx, pch=16, col=transparentColor('black', 0.2), cex=1.2)
    points(y1 ~ x1, col='dodgerblue', cex=1.2)
    lines(x1, locFit$b0 + locFit$b1*x1, col='black', lwd=2, lty=2)
    proportionalLabel(c(0, 0.14), rep(1.1, 2), text=FALSE, xpd=NA, type='l', lwd=2, lty=2)
    proportionalLabel(0.15, 1.1, substitute('Rank '*pos*': '*italic(y) == a~sy~b%.%italic(x), list(pos=rank, a=rounded(locFit$b0, 2), sy=ifelse(b1 < 0, ' - ', ' + '), b=rounded(abs(b1), 2))), xpd=NA, adj=c(0, 0.5))
    proportionalLabel(c(0, 0.14), rep(1.1, 2), text=FALSE, xpd=NA, type='l', lwd=2, lty=2)
    proportionalLabel(0.95, 0.95, paste0('n = ', length(y1)), xpd=NA, adj=c(1, 0.5), font=3, col='dodgerblue')

    # standardized residuals ~ x
    par(mai=c(0.6732, 0.5412, 0.5412, 0.2772), cex=0.8)
    yRange  <-  max(abs(c(floor(min(resids)), ceiling(max(resids)))))
    yRange  <-  c(-1*yRange, yRange)
    plot(resids ~ x1, xlab='Predictor', ylab='Std. residuals', xpd=NA, ylim=yRange, type='n', axes=FALSE)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    points(resids ~ x1, pch=16, col=transparentColor('dodgerblue', 0.5))
    abline(h=0, col=1, lwd=2)
    abline(h=c(-2, 2), lty=2)
    lf1  <-  loess(resids ~ x1)
    lines(x1, lf1$fitted, col='tomato', lwd=2)
    
    # standardized residuals ~ fitted values
    plot(resids ~ yHat, xlab='Fitted Values', ylab='Std. residuals', xpd=NA, ylim=yRange, type='n', axes=FALSE)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    points(resids ~ yHat, pch=16, col=transparentColor('dodgerblue', 0.5))
    abline(h=0, col=1, lwd=2)
    abline(h=c(-2, 2), lty=2)
    lf2  <-  loess(resids ~ yHat)
    lines(yHat, lf2$fitted, col='tomato', lwd=2)
    
    # qqnorm plot of standardized residuals
    par(mai=c(0.9732, 0.5412, 0.2412, 0.2772), cex=0.8)
    qqPlot  <-  qqnorm(resids, main='QQNorm plot of Std. Residuals', xpd=NA, plot=FALSE)
    plot(y1 ~ x1, data=qqPlot, xlab='Theoretical quantiles', ylab='Sample quantiles', xpd=NA, ylim=yRange, xlim=yRange, type='n', axes=FALSE)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    points(qqPlot$y ~ qqPlot$x, pch=16, col=transparentColor('dodgerblue', 0.5))
    qqline(resids, col='tomato')
    
    # histogram of standardized residuals
    histPlot  <-  hist(resids, breaks=20, plot=FALSE)
    plot(NA, xlab='Std. Residuals', ylab='Density', xpd=NA, ylim=c(0, max(histPlot$density)), xlim=yRange, type='n', axes=FALSE)
    usr  <-  par('usr')
    rect(usr[1], usr[3], usr[2], usr[4], col='grey90', border=NA)
    whiteGrid()
    box()
    axis(1, cex.axis=0.9)
    axis(2, las=1, cex.axis=0.9)
    densities  <-  histPlot$density
    breakPts   <-  histPlot$breaks
    for(j in seq_along(densities)) {
        polygon(c(breakPts[j], breakPts[j+1], breakPts[j+1], breakPts[j], breakPts[j]), c(rep(usr[3], 2), rep(densities[j], 2), usr[3]), border='dodgerblue', col=transparentColor('dodgerblue', 0.5))
    }
```

